
# Heat-Related Mortality

## Preparation

### Packages

```{r}
#| echo: true
#| results: false
#| warning: false
#| message: false

library(dlnm) # distributed lag non-linear models
library(lubridate) # date, time
library(mgcv) # flexible regressions
library(patchwork) # multiple graphics
library(renv) # R environment
library(rprojroot) # RProjekt root finden
library(splines) # non-linear models
library(tidyverse) # tidyverse

```


### Paths

```{r}
#| echo: true
#| results: false
#| warning: false
#| message: false

# project root
root <- find_root(is_rstudio_project)

# deaths: city of Zurich
path_death_zh <- "https://data.stadt-zuerich.ch/dataset/bev_tag_todesfaelle_quartier_geschl_ag_herkunft_od4211/download/BEV421OD4211.csv"

# deaths: FSO
path_death_fso <- file.path(root, "data/deaths-FSO.csv")

# file from Swiss Stats Explorer:

# https://dv.stats.swiss?chartId=25456e79-3ef6-4562-b23a-dad3cd67c139

# https://stats.swiss/vis?lc=de&fs[0]=Erhebung%252C%20Statistik%2C0%7CStatistik%20der%20nat%C3%BCrlichen%20Bev%C3%B6lkerungsbewegung%23BEVNAT%23&pg=10&fc=Erhebung%2C%20Statistik&bp=true&snb=19&df[ds]=ds%3Adisseminate&df[id]=DF_BEVNAT_EVENEMENTS_1&df[ag]=CH1.BEVNAT&df[vs]=1.0.0&dq=4.8100._T.D&to[TIME_PERIOD]=true&pd=1969-01-01%2C2025-12-31&vw=tb



# daily temperature at SMA Zurich (harmonized time series)
path_temp <- "https://data.geo.admin.ch/ch.meteoschweiz.ogd-nbcn/sma/ogd-nbcn_sma_d_historical.csv"



```


### Functions

```{r}
#| echo: true
#| results: false
#| warning: false
#| message: false

source(file.path(root, "R/RR_fun.R"))

```


### Parameter

```{r}

# last year with both Zurich and FSO data
# (complete year: January to December)

last_year <- 2025

```



### Graphics

```{r}
#| echo: true
#| results: false
#| warning: false
#| message: false

# colors
# https://github.com/StatistikStadtZuerich/zuericolors/blob/main/R/palettes.R

qual12 <- c("#3431DE", "#0A8DF6", "#23C3F1", "#7B4FB7", "#DB247D", "#FB737E",
            "#007C78", "#1F9E31", "#99C32E", "#9A5B01", "#FF720C", "#FBB900")


```


Grafiken mit Zoom-Funktionalität.

```{=html}
#| echo: true
#| results: false
#| warning: false
#| message: false

<script src = "https://ajax.googleapis.com/ajax/libs/jquery/3.4.1/jquery.min.js""></script>
<script type="text/javascript">
  $(document).ready(function() {
    $('body').prepend('<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>');
    // onClick function for all plots (img's)
    $('img:not(.zoomImg)').click(function() {
      $('.zoomImg').attr('src', $(this).attr('src')).css({width: '100%'});
      $('.zoomDiv').css({opacity: '1', width: 'auto', border: '1px solid white', borderRadius: '5px', position: 'fixed', top: '50%', left: '50%', marginRight: '-50%', transform: 'translate(-50%, -50%)', boxShadow: '0px 0px 50px #888888', zIndex: '50', overflow: 'auto', maxHeight: '100%'});
    });
    // onClick function for zoomImg
    $('img.zoomImg').click(function() {
      $('.zoomDiv').css({opacity: '0', width: '0%'});
    });
  });
</script>
```



## Import


### Deaths: FSO

```{r}

# import deaths
dea_imp <- read_csv(path_death_fso) |> 
  rename(date = TIME_PERIOD, deaths = OBS_VALUE) |> 
  select(date, deaths)

# duplicates? no
dea_imp |> 
  count(date) |> 
  filter(n == max(n))

# deaths for all days from May to September
dea <- tibble(date = seq(min(dea_imp$date), max(dea_imp$date), by = "day")) |> 
  left_join(dea_imp, by = "date") |> 
  mutate(deaths = replace_na(deaths, 0),
         month = month(date)) |>
  filter(month >= 5, month <= 9) |>
  select(date, month, deaths)


```

First plot of FSO death data


```{r}
#| label: fig-01
#| fig-cap: "deaths per day (FSO)"
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false

dea |> 
  ggplot() +
  geom_line(aes(x = date, y = deaths)) +
  facet_wrap(~as.factor(month), nrow = 1)

```

### Deaths: Zurich

```{r}

# import deaths
dea_imp_zh <- read_csv(path_death_zh) |> 
    mutate(date = as.Date(paste0(GueltigAbDatJahr, "-", GueltigAbDatMM, "-", GueltigAbDatDD))) |> 
    group_by(date) |>
    summarise(deaths = sum(AnzSterWir, na.rm = TRUE), .groups = "drop")

# duplicates? no
dea_imp_zh |> 
  count(date) |> 
  filter(n == max(n))

# deaths for all days from May to September
dea_zh <- tibble(date = seq(min(dea_imp_zh$date), max(dea_imp_zh$date), by = "day")) |> 
  left_join(dea_imp_zh, by = "date") |> 
  mutate(deaths = replace_na(deaths, 0),
         month = month(date)) |>
  filter(month >= 5, month <= 9) |>
  select(date, month, deaths)


```

```{r}
#| label: fig-02
#| fig-cap: "deaths per day (Zurich)"
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false

dea_zh |> 
  ggplot() +
  geom_line(aes(x = date, y = deaths)) +
  facet_wrap(~as.factor(month), nrow = 1)

```


### Temperature

```{r}

# import mean daily temperature
tem <- read_delim(path_temp, delim = ";") |>
  mutate(date = as.Date(reference_timestamp, format = "%d.%m.%Y %H:%M")) |>
  rename(temp = ths200d0) |>
  select(date, temp)

# checks
tem |>
  mutate(diff = c(1, diff(date))) |>
  summarize(min_date_diff = min(diff),
            max_date_diff = max(diff),
            temp_NA = sum(is.na(temp)))


```

First plot for temperature data


```{r}
#| label: fig-03
#| fig-cap: "mean temperature per day"
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false

tem |> 
  mutate(month = month(date)) |>
  filter(month >= 5, month <= 9) |> 
  ggplot() +
  geom_line(aes(x = date, y = temp)) +
  facet_wrap(~as.factor(month), nrow = 1)


```




### FSO-Death and temperature combined


```{r}

# data is limited by death values available
range(dea$date)
range(tem$date)

# imported data
total <- dea |>
  left_join(tem, by = "date")

# check
nrow(dea)
nrow(total)



```

## Analysis

### Relative risk (RR)

Relative risk estimation with FSO death data.

```{r}

# years (only if ten years with past data available)
years <- (year(min(total$date))+9) : year(max(total$date))

# data: relative risk
RR_dat <- map_dfr(years, function(y) {
  RR_fun(
    date = total$date,
    temp = total$temp,
    deaths = total$deaths,
    year_eval = y
  )
})

```


### Exposure-response relationship

Exposure-response relationship with FSO death data.

```{r}
#| label: fig-04
#| fig-cap: "exposure-response relationship"
#| fig-width: 15
#| fig-height: 12
#| echo: false
#| message: false
#| warning: false

ggplot(RR_dat, aes(x = temp, y = RR)) +
  geom_ribbon(aes(ymin = RR_low, ymax = RR_high), alpha = 0.2) +
  geom_line(linewidth = 1) +
  facet_wrap(~ period) +
  geom_hline(yintercept = 1) +
  geom_vline(aes(xintercept = MMT), linetype = "dotted") +
  labs(x = "mean daily temperature (°C)",
       y = "realtive risk") +
  theme_minimal()


```

### Attributable fraction

Attributable fraction with FSO death data.

```{r}

AF <- RR_dat |>
  filter(year_pred == 1) |>
  mutate(AF = pmax(0, if_else(temp > MMT, (RR - 1) / RR, 0)),
         AF_low = pmax(0, if_else(temp > MMT, (RR_low - 1) / RR_low, 0)),
         AF_high = pmax(0, if_else(temp > MMT, (RR_high - 1) / RR_high, 0)))
         


```

Cumulative distribution of the attributable fraction

```{r}
#| label: fig-05
#| fig-cap: "Cumulative distribution of attributable fraction"
#| fig-width: 6
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false

ggplot(AF, aes(x = AF)) +
  stat_ecdf(linewidth = 0.6) +
  labs(x = "attributable fraction", y = "F(x)") +
  theme_minimal()


```




```{r}
#| label: fig-06
#| fig-cap: "Cumulative distribution of attributable fraction by period"
#| fig-width: 10
#| fig-height: 5
#| echo: false
#| message: false
#| warning: false

ggplot(AF, aes(x = AF, col = period)) +
  stat_ecdf(linewidth = 0.3) +
  labs(x = "attributable fraction", y = "F(x)", col = "") +
  theme_minimal()


```



### Attributable deaths

Attributable deaths with Zurich death data.

Zurich death data begins later than the estimated AF; 
therefore, right join on Zurich death data.


```{r}

AD <- AF |> 
  select(year, date, temp, MMT, RR, RR_low, RR_high, period, 
         AF, AF_low, AF_high) |> 
  right_join(dea_zh, by = "date") |> 
  mutate(AD = AF * deaths,
         AD_low = AF_low * deaths,
         AD_high = AF_high * deaths) |> 
  filter(year <= last_year)


```


## Results

### Heat-related mortality by year

Heat-related mortality in the city of Zurich by year.

```{r}
#| label: fig-07
#| fig-cap: "heat-related mortality by year"
#| fig-width: 8
#| fig-height: 5
#| echo: false
#| message: false
#| warning: false

mort_y <- AD |>
  group_by(year) |>
  summarise(across(c(deaths, AD, AD_low, AD_high),
                   ~ sum(.x, na.rm = TRUE)),
            .groups = "drop")

ggplot(mort_y) +
  geom_col(aes(x = factor(year), y = AD), fill = qual12[6]) +
  geom_errorbar(aes(x = factor(year), ymin = AD_low, ymax = AD_high),
                width = 0.2, col = "grey40") +
  labs(x = "year", y = "heat-related deaths per year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

```

### Proportion of heat related deaths (summer months)

Proportion of heat related deaths from May to September per year (city of Zurich).

```{r}
#| label: fig-08
#| fig-cap: "attributable fraction by year"
#| fig-width: 8
#| fig-height: 5
#| echo: false
#| message: false
#| warning: false

prop_y_summer <- mort_y |> 
  mutate(percent = pmin(100, AD/deaths * 100),
         percent_low = pmin(100, AD_low / deaths * 100),
         percent_high = pmin(100, AD_high / deaths * 100))

ggplot(prop_y_summer) +
  geom_col(aes(x = factor(year), y = percent), fill = qual12[10]) +
  geom_errorbar(aes(x = factor(year), ymin = percent_low, ymax = percent_high),
                width = 0.2, col = "grey40") +
  labs(x = "year", y = "proportion of heat-related deaths (in %)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))


```

### Proportion of heat related deaths (entire year)

```{r}
#| label: fig-09
#| fig-cap: "proportion of heat related deaths (entire year)"
#| fig-width: 8
#| fig-height: 5
#| echo: false
#| message: false
#| warning: false

prop_y <- dea_imp_zh |>
  mutate(year = year(date)) |>
  filter(year >= min(mort_y$year), year <= last_year) |>
  group_by(year) |>
  summarize(total = sum(deaths), .groups = "drop") |>
  left_join(mort_y, by = "year") |>
  mutate(percent = pmin(100, AD / total * 100),
         percent_low = pmin(100, AD_low  / total * 100),
         percent_high = pmin(100, AD_high  / total * 100))

ggplot(prop_y) +
  geom_col(aes(x = factor(year), y = percent), fill = qual12[4]) +
  geom_errorbar(aes(x = factor(year), ymin = percent_low, ymax = percent_high),
                width = 0.2, col = "grey40") +
  labs(x = "year", y = "proportion of heat-related deaths (in %)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))


```





### heat-related mortality by year and heat intensity

In the city of Zurich.

```{r}
#| label: fig-10
#| fig-cap: "heat-related mortality by year and heat intensity"
#| fig-width: 8
#| fig-height: 5
#| echo: false
#| message: false
#| warning: false

# temperature classes
temp_classes <- c("moderate", "hot", "very hot")

with_temp_classes <- AD |>
  mutate(temp_class = factor(case_when(
    temp < 25 ~ temp_classes[1],
    (temp >= 25) & (temp < 27) ~ temp_classes[2],
    temp >= 27 ~ temp_classes[3]),
    levels = temp_classes)
  )

# mortality by year and temperature class
mort_yt <- with_temp_classes |>
  group_by(year, temp_class) |>
  summarise(AD = sum(AD), .groups = "drop")

ggplot() +
  geom_col(data = mort_yt,
           aes(x = factor(year), y = AD, fill = fct_rev(temp_class))) +
  geom_errorbar(data = mort_y,
                aes(x = factor(year), ymin = AD_low, ymax = AD_high),
                width = 0.2, col = "grey40") +
  scale_fill_manual(values = qual12[c(5, 11, 12)]) +
  labs(x = "year", y = "heat-related deaths per year", fill = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))



```


###  selected years (daily values)

```{r}
#| label: fig-11
#| fig-cap: "selected years (daily values)"
#| fig-width: 12
#| fig-height: 10
#| echo: false
#| message: false
#| warning: false


# Why these years? high values (2015, 2018), and last year (2025)
years_selected <- AD |>
  filter(year %in% c(2015, 2018, 2025)) |>
  mutate(plot_date = as.Date(paste0("2000-", format(date, "%m-%d"))))


p1 <- ggplot(years_selected) +
  geom_line(aes(x = plot_date, y = temp), col = qual12[1]) +
  geom_line(aes(x = plot_date, y = MMT), col = qual12[1], linetype = "dashed") +
  facet_wrap(~ factor(year)) +
  labs(x = "", y = "mean daily temperature (°C)") +
  theme_minimal()

p2 <- ggplot(years_selected) +
  geom_line(aes(x = plot_date, y = AD), col = qual12[5]) +
  geom_ribbon(aes(x = plot_date, ymin = AD_low, ymax = AD_high),
              fill = qual12[5], alpha = 0.2) +
  facet_wrap(~ factor(year)) +
  labs(x = "", y = "daily heat-related deaths") +
  theme_minimal()

p1 / p2


```

### MMT by year

```{r}
#| label: fig-12
#| fig-cap: "minimum mortality temperature (MMT) by year"
#| fig-width: 7
#| fig-height: 5
#| echo: false
#| message: false
#| warning: false

MMT_y <- AD |>
  group_by(year) |>
  summarise(MMT = unique(MMT), .groups = "drop")

ggplot(MMT_y) +
  geom_col(aes(x = factor(year), y = MMT), fill = qual12[1]) +
  labs(x = "year", y = "minimum mortality temperature") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))


```















...
